\thetitle

### 2021-02-22 writing/challenges.tex
stylistic patches, typo

The general challenge of evaluating blame is a methodological one. Unlike most
current research on programming languages, the question seems to call for
empirical research similar to that of human-computer interaction. At the same
time, a significant evaluation result demands a large number of cases yielding a
serious amount of data. As~\citet{lksfd-popl-2020} recently demonstrated, the
way around this dilemma is to postulate the equivalent of a {\it homo
economicus\/} for programming languages---the {\em rational programmer\/}---and
to implement this rational programmer as an algorithm.

In the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of debugging sessions yields data that
is similar to data collected in an empirical, human-facing manner.

The {\em first derivative challenge\/} is to find a representative collection of
impedance mismatches between typed and untyped pieces of code that programmers
may accidentally create and that the run-time checks of academic systems
catch. In other words, the experiment calls for a collection of type mistakes in
mix-typed programs that is representative of those ``in the wild.''
Unfortunately no such collection exists, and with good reason. The kind of
mistakes needed are typically detected by unit or integration tests; even if it
takes some time to find their sources, these mistakes do not make it into code
repositories with appropriate commit messages.

Following~\citet{lksfd-popl-2020} again, the alternative is to {\em generate a
sufficiently large and representative corpus of type mistakes \/} using mutation
analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis}.  The
mutation analysis of Lazarek et al, is, however, useless for creating impedance
mismatches.  While mutation analysis traditionally aims to inject bugs that
challenge test suites, it discards those that yield ill-typed mutants as
\emph{incompetent}.  Indeed, mutation analysis frameworks are fine-tuned to
avoid them---and yet it is precisely those that are needed for evaluating blame
assignment strategies.

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutatnt component.  A
debugging scenario results from removing the types from the mutatnt. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About two thirds of the original mutators
survived this first scrutiny testing. All of the remaining ones yield
interesting (see section~{sub:interesting}) debugging sessions.

@@

One way to overcome this challenge is to equip the rational programmer with
\emph{modes} that represent the alternative interpretations of blame.  For
instance, a mode of the rational programmer using Transient may pick the oldest
element of a blame set as the next component to add types because it corresponds
to the earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the blame set as another form of a
stack trace and add types to its newest element.  When both modes are equally
successful in locating an impedance mismatch, the method can measure the
rational programmer's effort to answer which mode is the most effective.

Different blame information is also due to differences in the run-time checks
that the three gradual typing systems perform. This observation suggests another
mode for the rational programmer, namely, a mode that ignores the blame
information completely and instead inspects only the information in the
exceptions of the safety checks in the underlying languages.  This
\emph{exception} mode serves as the baseline for blame's value; if the
programmer in this mode ``beats'' the others, blame systems might be useless.

The {\em third challenge\/} is the explosion of debugging scenarios that results
from the combination of mutation-based scenario generation and mode-based
analysis. All three factors---three different gradual typing systems, the large
number of mutants, and the number of debugging modes---contribute possibly
useful debugging data in multiplicative manner. Hence, carried out naively, the
experiment would demand an unacceptably large amount of computational resources.
A practical execution has no option but to sample the space of scenarios,
carefully ensuring that it still produces reproducible conclusions.

Overcoming these challenges needs a systematic approach: 
\begin{itemize}

\item a suitable set of mutators for the generation of interesting type-level
 errors (section~\ref{sec:mutate});

\item an algorithmic simulation of the many modes of the rational programmer on
 the corresponding debugging scenarios (section~\ref{sec:rational});

\item a careful sampling of the space (section~\ref{sec:sample}).
\end{itemize}
Getting all three aspects right is necessary to obtain a computationally
feasible and statistically sound experiment

{\bf Note} A somewhat coincidental challenge is the disparity of gradually typed
languages. While the three approaches to blame are well understood in theory,
they have been implemented for rather different untyped languages and with
vastly different type systems. To eliminate the effect of these differences, it
is necessary to implement all three approaches in a single language so that an
application of an evaluation method can use the same test programs, the same
type system, and hence the same debugging scenarios.

The results in this paper are for a system that builds on Greenman and
Felleisen's effort comparing the performance of Natural and Transient in
Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
Racket and comes with a Transient semantics that is comparable with the Natural
one. The Transient semantics of the system also realizes the design of the blame
assignment system proposed by~\cite{vss-popl-2017}. The Erasure semantics drops
the type information and runs the programs as if they were plain Racket
programs. The next section explains this system in detail.

>>>>

The general challenge of evaluating blame is a methodological one. Unlike most
current research on programming languages, the question seems to call for
empirical research similar to that of human-computer interaction. At the same
time, a significant evaluation result demands a large amount of
data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
dilemma is to postulate the {\em rational programmer\/}---the equivalent of a
{\it homo economicus\/} for programming languages---and to implement it as an algorithm.

In the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of debugging scenarios yields data that
is similar to data collected in an empirical, human-facing manner.

The {\em first derivative challenge\/} is to find a representative collection of
impedance mismatches. They must represent mistakes that programmers accidentally
create and that the run-time checks of academic systems catch. In other words,
the experiment calls for a collection of type mistakes in mix-typed programs
that is representative of those ``in the wild.''  Unfortunately no such
collection exists, and with good reason. The kind of mistakes needed are
typically detected by unit or integration tests; even if it takes some time to
find their sources, these mistakes do not make it into code repositories with
appropriate commit messages.

An alternative is to {\em generate a large and representative corpus of type
mistakes \/} using mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}.  Conventional mutation analysis is useless, however. Mutation
analysis traditionally aims to inject bugs that challenge test suites, and it
discards those that yield ill-typed mutants as \emph{incompetent}.  Indeed,
mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
precisely those that are needed for evaluating blame assignment strategies.

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutatnt component.  A
debugging scenario results from removing the types from the mutatnt. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About two thirds of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:interesting}).

@@

One way to overcome this challenge is to {\em equip the rational programmer with
modes \/} that represent the alternative interpretations of blame.  For
instance, one Transient mode may pick the oldest element of a blame set as
the next component to add types because it corresponds to the earliest point in
the program's evaluation that can discover an impedance mismatch.  Another mode
may opt to treat the sequence as a stack and add types to its newest element.
When both modes are equally successful in locating an impedance mismatch,
measuring the rational programmer's effort may answer which mode is the most
effective.

Different blame information is also due to differences in the run-time checks
that the three gradual typing systems perform. This observation suggests another
mode for the rational programmer, namely, a mode that ignores the blame
information completely and instead inspects only the information in the
exceptions of the safety checks in the underlying languages.  This exception
mode serves as the {\em baseline\/} for blame's value; if the programmer in this
mode ``beats'' the others, blame systems might be useless.

The {\em third challenge\/} is the explosive number of debugging scenarios that
result from the combination of mutation-based scenario generation and mode-based
analysis. All three factors---three different gradual typing systems, the large
number of mutants, and the number of debugging modes---contribute possibly
useful experimental data in a multiplicative manner. Hence, carried out naively,
the experiment would demand an unacceptably large amount of computational
resources.  A practical execution has no option but to {\em sample the space of
scenarios\/}, carefully ensuring that it still produces reproducible
conclusions.

Overcoming these challenges yields a systematic approach: 
\begin{itemize}

\item interesting mutators generate interesting type-level errors and thus
debugging scenarios (section~\ref{sec:mutate});

\item an algorithm simulates the modes of the rational programmer on these
scenarios (section~\ref{sec:rational}); and

\item a careful sampling of the space yields experimental
data about the bug finding process (section~\ref{sec:sample}).
\end{itemize}
Getting all three aspects right is necessary to obtain a computationally
feasible and statistically sound experiment.

{\bf Note} A coincidental challenge is the disparity of gradually typed
languages. While the three approaches to blame are well understood in theory,
they have been implemented for rather different untyped languages and with
vastly different type systems. To eliminate the effect of these differences, it
is necessary to implement all three approaches in a single language so that an
application of the evaluation method can use the same test programs, the same
type system, and hence the same debugging scenarios.

The results in this paper are for a system that builds on Greenman and
Felleisen's effort of comparing the performance of Natural and Transient in
Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
Racket and comes with a Transient semantics that faithfully implements the
design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
Erasure semantics drops the type information and runs the programs as if they
were plain Racket programs. The next section explains this system in detail.
